---
title: Analysis of U.S. National Oceanic and Atmospheric Administration's storm
  data
author: "Emmanuel ROCHE"
date: "08/03/2016"
output: html_document
---

# Synopsys

# Data Processing

The source dataset used in this analysis can be retrieved from , we keep the dataset in a compressed format since it can be read directly in R.

```{r}
if(!file.exists("StormData.csv.bz2"))
{
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "StormData.csv.bz2")
}


data <- read.csv("StormData.csv.bz2")
```

We can first check the number of observations available and display a few of them:

```{r}
dim(data)
```

This dataset contains `r dim(data[2])` variables, but for this initial analysis, we are only interested in a few of those variables:

  * **EVTTYPE** : indicates the type of the storm event that occured,
  * **LENGTH** : indicates the duration of the storm event,
  * **FATALITIES** : number of death due to the storm event,
  * **INJURIES** : number of injuries due to the storm event.

We isolate those variables in a new dataset, and we only consider the available complete cases.

```{r}

```

# Results

# Classification of the most 